Repeatability of Pre-Flashover Fire Patterns on Gypsum Wallboard

### 2017-03-19 Projects/Madrzykowski_Thesis/Report/Madrzykowski_Thesis.tex
Added front piece refs and working on discussion

\clearpage
Abbreviations
\begin{tabbing}
\hspace{1.5in} \= \\
EMS \> Emergency Medical Services \\
FDS \> Fire Dynamics Simulator \\
HGL \> Hot Gas Layer \\
HRR \> Heat Release Rate \\
HRRPUA \> Heat Release Rate per Unit Area \\
NFPA \> National Fire Protection Association  \\
NIST \> National Institute of Standards and Technology \\
\end{tabbing}

@@

@@

In addition a few popular numerical models were to use simulate the fires and the resulting fire patterns based on input data collected from engineering reference sources, the bench-scale, and the full-scale fire experiments conducted in this study.

@@

@@

@@

The main objective of the research was to examine the repeatability of pre-flashover fire patterns generated from exposure to short duration (300 s maximum) well characterized fires.  The first step was to examine the repeatability of different pre-flashover fires.  Three different fuels were used: natural gas, gasoline, and polyurethane foam.  Each fuel had a similar top surface area.  The heat release rate data showed that the variability was greater for more complex fuels.  While the variation in peak heat release rate with the natural gas was similar to the expanded uncertainty of the measurement system, $\pm$ 11~\%, the variation in the peak heat release rates of the gasoline and the polyurethane foam increased due to increased uncertainties in the burning behavior of the fuels.  The fire pattern heights generated on painted gypsum wallboard by the natural gas burner and gasoline fueled fires were shown to have uncertainties of $\pm$ 18~\% or less based on a Type A statistical analysis with 95~\% confidence limits. The patterns generated by the polyurethane foam fueled fire, had more variability than the natural gas and the gasoline fire patterns.  This is likely due to the greater variability of the burning of the solid fuel and the lower heat release rate.    The comparison of the fire pattern heights and the mean flame height demonstrated that the “steady state” natural gas fires exhibited the highest level of agreement and polyurethane foam fueled fire exhibited the least agreement, with the gasoline fueled fires in between.  This trend is the same as exhibited by the repeatability of the heat release rate measurements.  Given that heat release rate is the source of the energy which is transferred to the walls in order generate the fire patterns, it stands to reason that the greater the uncertainty in the heat release rate the greater the uncertainty in pattern and any associated thermal measurements.

>>>>

@@

\textbf{Material covered in Chapters 2 and 6 has been previously published in:}

Madrzykowski, D. and Fleischmann, C., Fire Pattern Repeatability: A Study in Uncertainty, Journal of
Testing and Evaluation, ASTM International, W. Conshohocken, PA.,Vol 40, No. 1, January 2012, pp 1-11.

Madrzykowski, D. and Fleischmann, C., "Fire Pattern Repeatability: A Laboratory Study on Gypsum Wallboard." In Proceedings ISFI 2010: International Symposium on Fire Investigation Science and Technology, Hyattsville, MD, USA. Sarasota, FL, USA 2010. pp 381-392.

\textbf{Material covered in Chapters 2 and 7 has been previously published in:}

Madrzykowski, D. and Fleischmann, C., "Flame Heights and Heat Transfer from Gaseous, Liquid and Solid Fuels for Burn Pattern Repeatability Research." In INTERFLAM 2010: 12th International Conference on Fire Science and Engineering, Nottingham, UK. London, UK, 2010. pp 1930-1942.

@@

In addition a few popular numerical models were used to simulate the fires or the effects of the fires and the resulting fire patterns based on input data collected from engineering reference sources, the bench-scale experiments, and the full-scale fire experiments conducted as part of this study.  The results of the calculations were then compared to the measured values to examine the level of agreement.

In Chapters 2 through 8 results from specific test series or calculations were presented.  Typically there was some discussion specific to the experiments or calculations being addressed in each chapter.  In this chapter the discussion will be more inclusive in examining the key outcomes and the remaining gaps.

\section{Repeatability of Pre-Flashover Fire Patterns}

As noted in the Introduction, gypsum wall board was chosen as a target for the fire patterns due to its common usage in construction and the frequency of fires that occur in buildings where gypsum wall board may be used as an interior finish. The size of the fires were chosen such that the fire damaged area (paper consumed) of the pattern was contained to less than 2.4~m.

@@

`Essentially, all models are wrong, but some are useful.'   Quote by George E.P. Box  statistician

@@

Uncertainty and Modeling

NFPA 921, Guide for Fire and Explosion Investigations considers mathematical modeling techniques as a means to provide the fire investigator with tools for testing hypotheses regarding the origin and cause of the fire and the cause of the resulting damage to property or injury to people.  NFPA 921 cautions that `mathematical modeling, whether simplified hand calculations or computer fire models, has inherent limitations and assumptions that should be considered. Models generally rely upon empirical data and are validated via comparison with other empirical data. Care must be taken to assure that the model is being used with due regard for limitations, assumptions, and validation'~\cite{NFPA:921}.  In these experiments, the natural gas burner and the gasoline pool fire were the type of fire source that the models were developed with or have been validated against.  Basically the burners provided fire sources that were very repeatable and could generally be defined by the two dimensional planar surface where the fuel is emitted.

Fire investigators, however, examine fire scenes where the first item ignited may have involved a piece of furniture, an appliance, or even a car.   These items are three dimensional and the fuel could be emitted from multiple surfaces at any given time in the fire event.  Further thermoplastics are commonly used in the construction of these items, with furniture being a prevalent item.  As the thermoplastic is heated it melts, flows, drips, and burns.   As a result the flame geometry generated by a piece of furniture may not be representative of the flame geometry characterized by the studies using gas burners and pool fires.  In this study even using a block of polyurethane foam with a simple geometry led to increased uncertainty.

@@

The main objective of the research was to examine the repeatability of pre-flashover fire patterns generated from exposure to short duration (300 s maximum) well characterized fires.  The first step was to examine the repeatability of different fire sources.  Three different fuels were used: natural gas, gasoline, and polyurethane foam.  Each fuel had a similar top surface area.  The heat release rate data showed that the variability was greater for more complex fuels.  While the variation in peak heat release rate with the natural gas was similar to the expanded uncertainty of the measurement system, $\pm$ 11~\%, the variation in the peak heat release rates of the gasoline and the polyurethane foam increased due to increased uncertainties in the burning behavior of the fuels.  The fire pattern heights generated on painted gypsum wall board by the natural gas burner and gasoline fueled fires were shown to have uncertainties of $\pm$ 18~\% or less based on a Type A statistical analysis with 95~\% confidence limits. The patterns generated by the polyurethane foam fueled fire, had more variability than the natural gas and the gasoline fire patterns.  This is likely due to the greater variability of the burning of the solid fuel and the lower heat release rate.    The comparison of the fire pattern heights and the mean flame height demonstrated that the “steady state” natural gas fires exhibited the highest level of agreement and polyurethane foam fueled fire exhibited the least agreement, with the gasoline fueled fires in between.  This trend was the same as exhibited by the repeatability of the heat release rate measurements.  Given that heat release rate is the source of the energy which was transferred to the walls in order generate the fire patterns, it stands to reason that the greater the uncertainty in the heat release rate the greater the uncertainty in pattern and any associated thermal measurements.
